FHJ: A Formal Model for Hierarchical Dispatching and Overriding

### 2017-10-17 NewVersion/sections/Introduction.tex
Abstract intro overview

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. It is that the conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{triangle inheritance}, in
analogy to diamond inheritance.

@@

\begin{figure}
\center  \includegraphics[height=4cm]{pics/DrawableSafeDeck3.pdf}
\caption{\lstinline|DrawableSafeDeck|: an illustration of hierarchical overriding.}
\label{fig:main}
\end{figure}

\textit{Hierarchical overriding} is a novel language
mechanism that allows method overriding to be applied
only to one branch of the class hierarchy. Hierarchical overriding 
adds expressive power that is not available in languages such as 
C++ or C\#. Hierarchical overriding allows overriding to work 
for classes with multiple (conflicting) methods sharing the 
same names and signatures. An example is presented in 
Figure~\ref{fig:main}. In this example, there are 4 classes/interfaces.
Two classes model a deck of cards (\lstinline|Deck|) and a class for
drawable widgets (\lstinline|Drawable|). 
The class \lstinline|SafeDeck| adds functionality to check whether the
deck is empty and to prevent drawing a card from an empty deck. 
The interesting class is \lstinline|DrawableSafeDeck|, which inherits 
from both \lstinline|SafeDeck| and \lstinline|Drawable|. Hierarchical
overriding is used in \lstinline|DrawableSafeDeck| to keep two separate \lstinline|draw|
methods for each parent, but overriding \emph{only} the \lstinline|draw|
method coming from \lstinline|Drawable|, in order to draw a widget
with a deck of cards. Note that hierarchical overriding is denoted in the UML
diagram with the notation
\lstinline|draw()|$\shneg$\lstinline|Drawable|. It expresses that the \lstinline|draw| method from \lstinline|Drawable|
is overridden. Although in this example only one of the
\lstinline|draw| methods is overridden (and the other is simply inherited), hierarchical overriding
supports multiple conflicting methods to be independently overridden as well.

To present hierarchical overriding and dispatching, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. 
In summary, our contributions are:
\begin{itemize}
	\item \textbf{Formalization of a hierarchical dispatching algorithm} that integrates both the static type and dynamic type for method dispatch, and hence
	ensures unambiguity as well as extensibility in the presence
        of unintentional method conflicts.
	\item \textbf{Hierarchical overriding:} a novel notion that allows
          methods to override individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on
          Featherweight Java, supporting the above features. 
          We provide the static and dynamic semantics and prove the
          type soundness of the model.
	\item \textbf{Prototype implementation\footnote{Available in
              supplementary material attached to the submission.}:} a
          simple implementation of \MIM{} interpreter in Scala.
\end{itemize}

>>>>

In contrast to diamond inheritance, the second case of ambiguity
is \textit{unintentional method conflicts}~\cite{scharli03traits}. In
such a case conflicting 
methods do not actually refer to the same feature. 
In a nominal system, methods can be designed for different
functionality, but happen to have the same names (and signatures).
A simple example of this situation is two \lstinline{draw} methods that
are inherited from a deck of cards and a drawable widget. 
In such context, the two \lstinline{draw} methods have very different meanings, 
but they happen to share the same name.
%%This issue was proposed by the trait paper, so-called
When inheritance is used to compose these methods, a compilation 
error happens due to conflicts. However, unlike the diamond problem,
the conflicting methods have very different meanings and do not share a
common parent. We call such a case \textit{triangle inheritance}, in
analogy to diamond inheritance.

@@

\begin{figure}[t]
\center  \includegraphics[height=4cm]{pics/DrawableSafeDeck3.pdf}
\caption{\lstinline|DrawableSafeDeck|: an illustration of hierarchical overriding.}
\label{fig:main}
\end{figure}

\textit{Hierarchical overriding} is a novel language
mechanism that allows method overriding to be applied
only to one branch of the class hierarchy. Hierarchical overriding 
adds expressive power that is not available in languages such as 
C++ or C\#. Hierarchical overriding allows overriding to work 
for classes with multiple (conflicting) methods sharing the 
same names and signatures. An example is presented in 
Figure~\ref{fig:main}. In this example, there are 4 classes/interfaces.
Two classes model a deck of cards (\lstinline|Deck|) and a class for
drawable widgets (\lstinline|Drawable|). 
The class \lstinline|SafeDeck| adds functionality to check whether the
deck is empty and to prevent drawing a card from an empty deck. 
The interesting class is \lstinline|DrawableSafeDeck|, which inherits 
from both \lstinline|SafeDeck| and \lstinline|Drawable|. Hierarchical
overriding is used in \lstinline|DrawableSafeDeck| to keep two separate \lstinline|draw|
methods for each parent, but overriding \emph{only} the \lstinline|draw|
method coming from \lstinline|Drawable|, in order to draw a widget
with a deck of cards. Note that hierarchical overriding is denoted in the UML
diagram with the notation
\lstinline|draw()|$\shneg$\lstinline|Drawable|, expressing that the \lstinline|draw| method from \lstinline|Drawable|
is overridden. Although in this example only one of the
\lstinline|draw| methods is overridden (and the other is simply inherited), hierarchical overriding
supports multiple conflicting methods to be independently overridden as well.

To present hierarchical overriding and dispatching, we introduce a
formalized model \MIM{} in Section~\ref{sec:formalization} based on
Featherweight Java~\cite{Igarashi01FJ}, together with theorems and
proofs for type soundness. 
In summary, our contributions are:
\begin{itemize}
	\item \textbf{A formalization of a hierarchical dispatching algorithm} that integrates both the static type and dynamic type for method dispatch, and 
	ensures unambiguity as well as extensibility in the presence
        of unintentional method conflicts.
	\item \textbf{Hierarchical overriding:} a novel notion that allows
          methods to override individual branches of the class hierarchy.
	\item \textbf{\name:} a formalized model based on
          Featherweight Java, supporting the above features. 
          We provide the static and dynamic semantics and prove the
          type soundness of the model.
	\item \textbf{Prototype implementation\footnote{\bruno{URL
                here} URL}:} a
          simple implementation of \MIM{} interpreter in Scala.
\end{itemize}

### 2017-10-17 NewVersion/sections/Overview.tex
Abstract intro overview

\section{A Running Example: Drawable Deck}~\label{sec:overview}
This section illustrates the problem of unintentional method conflicts,
together with the features of our model on addressing this issue, by
a simple running example. In the following text we will introduce three
problems one by one and have a discussion on possible workarounds and
our solutions.
Problems 1 and 2 are related to hierarchical dispatching, and in 
C++ it is possible to have similar solutions as ours to both
problems. Hence it is important to emphasize that, with respect to
hierarchical dispatching, our contribution is not a novel
mechanism. Instead, 
inspired by the C++ solution, our contribution is formalizing a minimal calculus
of this feature together with a proof of type soundness. However, for the final problem, there is no satisfactory approach
in existing languages, thus we propose a new feature (hierarchical
overriding) with the corresponding formalization of that feature.

@@

interface Drawable {
  void draw() { // draws something on the screen
    JFrame frame = new JFrame();
    frame.setVisible(true);
  }
}
\end{lstlisting}\vspace{3pt}
In \lstinline|Deck|,
\lstinline|draw| uses \lstinline|println|, which is a
library function. 
The two \lstinline|draw| methods can have different return types, but for simplicity, 
they both return \lstinline|void| here. Note that, similarly to
Featherweight Java~\cite{Igarashi01FJ}, \lstinline|void| is
unsupported in our formalization. We could have also defined an interface called \lstinline|Void|
and return an object of that type instead. To be concise, we use \lstinline|void| in this code.
For \lstinline|Drawable| the \lstinline|draw|
creates a blank canvas by \lstinline|JFrame| and set the visibility to \lstinline|true|.

@@

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is for composing the features from various 
components and to achieve code reuse, as supported by many mainstream OO
languages. Nevertheless at this point, languages like Java simply treat the two \lstinline|draw| methods
as the same, and hence the compiler throws an error
on \lstinline|DrawableDeck| about the conflict.

@@

\paragraph{\MIM{}'s solution:} To solve this problem it is important to preserve both conflicting methods
during inheritance instead of merging them into a single
method. Therefore \MIM{} accepts the definition of \lstinline|DrawableDeck|. To disambiguate method calls, 
we can use \emph{up-casts} in \MIM{}, so as to specify the ``branch'' in the
inheritance hierarchy that should be called. The following code
illustrates the use of up-casts for disambiguation:

@@

\vspace{3pt}\begin{lstlisting}
interface SafeDeck extends Deck {
  boolean isEmpty() {...}
  void draw() { // overriding
    if (isEmpty()) println("The deck is empty.");
    else println("Draw a card");
  }
}
\end{lstlisting}\vspace{3pt}
Without dynamic dispatching, we may have to copy the \lstinline|shuffleAndDraw| code into \lstinline|SafeDeck|, so that \lstinline|shuffleAndDraw| calls the new \lstinline|draw| defined in \lstinline|SafeDeck|. Yet dynamic dispatching immediately saves us from the duplication work,
since the method becomes automatically dispatched to the most specific one. Nevertheless, as seen before, dynamic dispatch would potentially introduce ambiguity,
for instance, when we have the class hierarchy structure shown in Figure~\ref{fig:drawablesafedeck}(left) and the following code:

@@

\begin{figure*}[t]
  % \nocaptionrule
  \centering
  \begin{minipage}[t]{0.32\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck1.pdf}
  \end{minipage}
  \centering
  \hspace*{2pt}
  \begin{minipage}[t]{0.38\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck0.pdf}
  \end{minipage}
  \centering
  \hspace*{2pt}
  \begin{minipage}[t]{0.25\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck3.pdf}
  \end{minipage}  
  \caption{UML graphs for \lstinline|DrawableSafeDeck|.}\label{fig:drawablesafedeck}
\end{figure*}

@@

// main program
new DrawableSafeDeck().shuffleAndDraw() // SafeDeck.draw is called
\end{lstlisting}\vspace{3pt}
During the runtime, when \lstinline|this.draw()| in called,
the dynamic type of the receiver is \lstinline|DrawableSafeDeck|, but the static type is actually
\lstinline|Deck|. Here it is an instance of the \textit{hierarchical invocation}, which can be read as 
``finding the most specific \lstinline|draw| along the path \lstinline|Deck|''. The meaning of ``along path
\lstinline|Deck|'' not only implies what \lstinline|Deck| inherits, but also the domain of its subtypes up to the
dynamic type \lstinline|DrawableSafeDeck|. Finally \lstinline|SafeDeck.draw| is correctly returned.

%Note that in C++, if we put the \lstinline[language=c++]|virtual| keyword in front of every \lstinline|draw| method, that means
%we enforce the dynamic dispatch, and it does the disambiguation correctly.
%Unfortunately, the C++ language, or rather the feature on triangle inheritance, has not been well formalized.
In our language model, there are two types of casts: \emph{implicit casts} and \emph{explicit upcasts}. The implicit casts appear
when passing arguments and returning expressions, they are also
upcasts. Hence all casts are safe.

@@

The UML graph is shown in Figure~\ref{fig:drawablesafedeck}(right), where
the up-arrow $\uparrow$ is short for ``\lstinline|override|''.
Here the idea is that \emph{only} \lstinline{Drawable.draw} is
overridden. This is accomplished by specifying, in the method
definition, that the method only overrides the \lstinline|draw| 
from \lstinline|Drawable|.  
The individual overriding allows us to make use of the methods from \lstinline|SafeDeck| as well. In the formalization, the 
hierarchical overriding feature is also an important feature that involves in the algorithm of hierarchical dispatch.
%\bruno{the last two sentences are a bit unclear. I think it is related to ``interaction''.}

@@

Before the formalized algorithm, Figure~\ref{fig:examplesmbody} gives a peek at the behavior by
some quick examples. The UML diagrams present the hierarchy. 
In (d) and (e), a cross mark indicates that the interface fails to type-check. Generally, \MIM{} rejects on the definition of an interface
during compilation, if it reaches a diamond and hence is possible to cause ambiguity. $\mbody$ is the method lookup function for hierarchical
dispatch, formally defined in Section~\ref{subsec:mbodydef}. In general, $\mbody(m, X, Y) = (Z, ...)$ reflects that the source code \lstinline|((X) new Y()).m()| calls \lstinline|Z.m| in the runtime. It is undefined when method dispatch is ambiguous.

>>>>

\section{A Running Example: Drawable Deck}~\label{sec:overview}
This section illustrates the problem of unintentional method conflicts,
together with the features of our model on addressing this issue, by
a simple running example. In the following text we will introduce three
problems one by one and have a discussion on possible workarounds and
our solutions.
Problems 1 and 2 are related to hierarchical dispatching, and in 
C++ it is possible to have similar solutions to both
problems. Hence it is important to emphasize that, with respect to
hierarchical dispatching, our contribution is not a novel
mechanism. Instead, 
inspired by the C++ solution, our contribution is formalizing a minimal calculus
of this feature together with a proof of type soundness. However, for the final problem, there is no satisfactory approach
in existing languages, thus what we propose is a new feature (hierarchical
overriding) with the corresponding formalization of that feature.

@@

interface Drawable {
  void draw() { // draws something on the screen
    JFrame frame = new JFrame();
    frame.setVisible(true);
  }
}
\end{lstlisting}\vspace{3pt}
In \lstinline|Deck|,
\lstinline|draw| uses \lstinline|println|, which is a
library function. 
The two \lstinline|draw| methods can have different return types, but for simplicity, 
they both return \lstinline|void| here. Note that, similarly to
Featherweight Java~\cite{Igarashi01FJ}, \lstinline|void| is
unsupported in our formalization. We could have also defined an interface called \lstinline|Void|
and return an object of that type instead. To be concise, however, we
use \lstinline|void| in our examples.
For \lstinline|Drawable| the \lstinline|draw| method
creates a blank canvas.

@@

\vspace{3pt}\begin{lstlisting}
interface DrawableDeck extends Drawable, Deck {} 
\end{lstlisting}\vspace{3pt}
The point of using multiple inheritance is for composing the features from various 
components and to achieve code reuse, as supported by many mainstream OO
languages. Nevertheless, at this point, languages like Java simply treat the two \lstinline|draw| methods
as the same, and hence the compiler throws an error
on \lstinline|DrawableDeck| about the conflict.

@@

\paragraph{\MIM{}'s solution:} To solve this problem it is important to preserve both conflicting methods
during inheritance instead of merging them into a single
method. Therefore \MIM{} accepts the definition of
\lstinline|DrawableDeck|\bruno{Marco complained about this sentence,
  which is in the todo list. I've then complained that we should have
  the todos at the places where TODOs were meant. Now, the sentence
  remains unchanged. PLEASE go over the TODO list, and check whether
  you've fixed the issue or not, if not move the TODO to the correct
  location! }. To disambiguate method calls, 
we can use \emph{up-casts} in \MIM{} to specify the ``branch'' in the
inheritance hierarchy that should be called. The following code
illustrates the use of up-casts for disambiguation:

@@

\vspace{3pt}\begin{lstlisting}
interface SafeDeck extends Deck {
  boolean isEmpty() {...}
  void draw() { // overriding
    if (isEmpty()) println("The deck is empty.");
    else println("Draw a card");
  }
}
\end{lstlisting}\vspace{3pt}
Without dynamic dispatching, we may have to copy the \lstinline|shuffleAndDraw| code into \lstinline|SafeDeck|, so that \lstinline|shuffleAndDraw| calls the new \lstinline|draw| defined in \lstinline|SafeDeck|. Yet dynamic dispatching immediately saves us from the duplication work,
since the method becomes automatically dispatched to the most specific one. Nevertheless, as seen before, dynamic dispatch would potentially introduce ambiguity.
For instance, when we have the class hierarchy structure shown in Figure~\ref{fig:drawablesafedeck}(left) and the following code:

@@

\begin{figure*}[t]
  % \nocaptionrule
  \centering
  \begin{minipage}[t]{0.32\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck1.pdf}
  \end{minipage}
  \centering
  \hspace*{2pt}
  \begin{minipage}[t]{0.38\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck0.pdf}
  \end{minipage}
  \centering
  \hspace*{2pt}
  \begin{minipage}[t]{0.25\textwidth}
  \includegraphics[height=4cm]{pics/DrawableSafeDeck3.pdf}
  \end{minipage}  
  \caption{UML diagrams for 3 variants of \lstinline|DrawableSafeDeck|.}\label{fig:drawablesafedeck}
\end{figure*}

@@

// main program
new DrawableSafeDeck().shuffleAndDraw() // SafeDeck.draw is called
\end{lstlisting}\vspace{3pt}
\bruno{Marco complined about this but this was not addressed either!!!}
During the runtime, when \lstinline|this.draw()| in called,
the dynamic type of the receiver is \lstinline|DrawableSafeDeck|, but the static type is actually
\lstinline|Deck|. Here it is an instance of the \textit{hierarchical invocation}, which can be read as 
``finding the most specific \lstinline|draw| along the path \lstinline|Deck|''. The meaning of ``along path
\lstinline|Deck|'' not only implies what \lstinline|Deck| inherits, but also the domain of its subtypes up to the
dynamic type \lstinline|DrawableSafeDeck|. Finally \lstinline|SafeDeck.draw| is correctly returned.

%Note that in C++, if we put the \lstinline[language=c++]|virtual| keyword in front of every \lstinline|draw| method, that means
%we enforce the dynamic dispatch, and it does the disambiguation correctly.
%Unfortunately, the C++ language, or rather the feature on triangle inheritance, has not been well formalized.
In our language model, there are two types of casts: \emph{implicit casts} and \emph{explicit upcasts}. The implicit casts appear
when passing arguments and returning expressions, they are also
upcasts. Hence all casts are safe. \bruno{This discussion seems
  inconplete. Either complete it or remove it.}

@@

The UML graph is shown in Figure~\ref{fig:drawablesafedeck}(right), where
the up-arrow $\uparrow$ is short for ``\lstinline|override|''.
Here the idea is that \emph{only} \lstinline{Drawable.draw} is
overridden. This is accomplished by specifying, in the method
definition, that the method only overrides the \lstinline|draw| 
from \lstinline|Drawable|.  
The individual overriding allows us to make use of the methods from \lstinline|SafeDeck| as well. In the formalization, the 
hierarchical overriding feature is also an important feature, involved in the algorithm of hierarchical dispatch.
%\bruno{the last two sentences are a bit unclear. I think it is related to ``interaction''.}

@@

Before the formalized algorithm, Figure~\ref{fig:examplesmbody} gives
a peek at the behavior 
using a few examples. The UML diagrams present the hierarchy. 
In (d) and (e), a cross mark indicates that the interface fails to type-check. Generally, \MIM{} rejects on the definition of an interface
during compilation, if it reaches a diamond and hence is possible to cause ambiguity. $\mbody$ is the method lookup function for hierarchical
dispatch, formally defined in Section~\ref{subsec:mbodydef}. In general, $\mbody(m, X, Y) = (Z, ...)$ reflects that the source code \lstinline|((X) new Y()).m()| calls \lstinline|Z.m| in the runtime. It is undefined when method dispatch is ambiguous.
